Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus

### 2024-01-30 manuscript_thev_transcriptome.Rmd
added plot for SSites; added to discussion

Myriads of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs -- specifically human AdVs. Thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV genes are transcribed in a temporal manner; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late unit (MLTU or major late promoter [MLP] region), which generates five families of late mRNAs (L1-L5) based on the polyadenylation site. An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host cell modulating proteins, primarily involved in DNA replication or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins that act as capsid proteins, promote virion assembly, and direct genome packaging. The immediate early gene E1A is expressed first, followed by the the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. Noteworthily, the MLP shows basal transcriptional activity during early infection (before DNA replication), with a comparable efficiency to other early viral promoters, but reaches its maximal activity during late infection (after DNA replication). However, during early infection the repertoire of late transcripts from the MLP is restricted until late infection [@Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. For instance, the MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader, TPL) [@Donovan2020; @Zhao2014]. There is also an alternate 5' three non-coding exons present in varying amounts on a subset of MLTU mRNAs (known as the x-, y- and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5) based on polyadenylation site. Each transcription unit (TU) contains its own promoter driving the expression of all the array of mRNA transcripts produced via alternative splicing in the unit [@Zhao2014; @Donovan2020; @Guimet2016]. The promoters are activated at different phases of the infection by proteins from previously activated TUs. Paradoxically, the early-to-late phase transition during infection requires the L4 genes, 22K and 33K, which should only be available after the transition. However, a promoter in the L4 region (L4P) that directs the expression of these two proteins independent of the MLP was found, resolving the paradox [@Morris2010; @Guimet2016; @Donovan2020]. During translation of AdV mRNA, recent studies strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020].

High throughput sequencing methods have facilitated the discovery of many novel transcribed regions and splicing isoforms. It is also a very powerful tool to study alternative splicing under different conditions at an unparalleled depth [@Djebali2012; @Zhao2014]. In this paper, a paired-end deep sequencing experiment was performed to characterize for the first time the transcriptome of THEV (VAS vaccine strain) during different phases of the infection, yielding the first THEV splicing map. Our paired-end sequencing allowed for reading `r seq_len` bp long high quality (mean Phred Score of 36) sequences from each end of cDNA fragments, which were mapped to the genome of THEV.
\newpage

@@

**Overview of sequencing data and analysis pipeline outputs**  
A previous study by Zeinab *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*Meleagris gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (**Table 1**, **Figure 2a**). From the mapped reads, we identified a total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junction and mapping reads to the THEV genome over time denotes an active infection, and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (**Figure 2b**).

@@

We validated the splice junctions in all transcripts by PCR amplification of viral cDNA, cloning, and Sanger sequencing (**Supplementary PCR methods**). During validation, we identified 5 additional transcripts some of which were further validated by 3' Rapid Amplification of cDNA Ends (3'RACE) data. The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (<http://www.ncbi.nlm.nih.gov/geo>) under **accession no. XXXXXX**.

@@

We also analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. We found that most splice donor-acceptor sequences were unsurprisingly the canonical GU-AG nucleotides.

@@

Only two ORFs (ORF1 [sialidase] and Hyd) are predicted in this region; however, we discovered `r as.character(english(nrow(filter(spliced_gtf, region == "E1"))))` novel transcripts in this region, which collectively contain `r nrow(reg_e1_brkdown)` unique splice junctions (**Figure 5**). Most of the ORFs of the novel transcripts are distinct from the predicted ORFs, but they all have the coding potential (CP) for the predicted Hyd protein as the 3'-most coding sequence (CDS) if secondary start codon usage is considered as reported for other AdVs [@Donovan2020; @Zhao2014]. The 5'-most CDS of TRXPT_1 is multi-exonic, encoding a novel 17.9 kilodalton (kDa), 160 residue [amino acids (aa)] protein (ORF9). From its 5'-most start codon (SSC), TRXPT_2 encodes the largest protein in this region -- a 64.3 kDa, 580 aa protein (ORF10) with the same SSC as ORF9 (position 211bp). ORF10 spans almost the entire predicted ORF1 and Hyd, coming short in two regards: it is spliced from 1655bp to 1964bp (ORF1's C-terminus, including the stop codon), and it's stop codon (STC; position 2312) is 13 bp short of Hyd's STC. However, it has an SSC 102 bp upstream and in-frame with ORF1's predicted SSC. Thus, ORF10 shares substantial protein sequence similarity with ORF1 but not with Hyd, as the SSC of Hyd is not in-frame. Without its splice site removing the ORF1 STC, TRXPT_2 would encode a longer variant of ORF1, starting from an upstream SSC. TRXPT_3 is almost identical to TRXPT_1, except for the lack of TRXPT_1's second exon. Our RNA-seq data shows that all E1 transcripts share the same transcription termination site (TTS; at position 2325bp). However, TRXPT_3 and TRXPT_4 seem to have transcription start sites (TSS) downstream of the TSS of TRXPT_1 and TRXPT_2 (E1 TSS; position: 54bp). Given that studies in MAdVs show that E1 mRNAs share not only a common TTS but also the TSS, and only differ from each other regarding the internal splicing [@Zhao2014], it is likely that TRXPT_3 and TRXPT_4 are incomplete, and their actual TSS just like the TTS are identical for all E1 transcripts. Regardless of the TSS considered for TRXPT_3, the coding potential (CP) remains unaffected. Its 5'-most CDS, beginning at 1965bp and sharing the same STC as ORF9, produces a 13.1 kDa, 115 residue protein (ORF4). ORF4 was predicted in an earlier study [@Pitcovski1998] but was excluded in later studies [@Beach2009a; @Davison2003]; however, our data suggests it is a bona fide ORF. Unlike TRXPT_3, the CP of TRXPT_4 is affected by the TSS considered; if we consider its unmodified TSS, then its CP is the same as TRXPT_3 (ORF4 as the first CDS and Hyd as second CDS using the secondary SSC). However, if we assume that TRXPT_4 uses the E1 TSS, then the 5'-most CDS is a distinct, novel, multi-exonic 15.9 kDa, 143 aa protein (ORF11) with the same SSC as ORF9 and ORF10 but with a unique STC. The splice junctions of all transcripts in this region (except the junction for TRXPT_4) were validated by cloning of viral cDNA and Sanger sequencing (**Supplementary PCR methods**).

@@

**Early Region 2 (E2) and Intermediate Region (IM) transcripts**     
The E2 TU expressed on the anti-sense strand is subdivided into E2A and E2B and encodes three classical AdV proteins --  pTP and Ad-pol (E2B proteins), and DBP (E2A protein) -- essential for genome replication [@Donovan2020; @Zhao2014]. Unlike MAdV where two promoters (E2-early and E2-late) are known [@Donovan2020], we discovered only a single TSS (E2 TSS; 18,751bp) from which both E2A and E2B transcription is initiated. However, similar to MAdVs, E2A and E2B transcripts have distinct TTSs, and the E2B transcripts share the TTS of the IVa2 transcript of the IM region [@Donovan2020; @Zhao2014] (**Figure 6**).

The E2A ORF, DBP is one of three THEV ORFs predicted to be spliced from two exons. The corresponding transcript (TRXPT_21) found in our data matches this predicted splice junction precisely but with a non-coding additional exon at the 5'-end (E2-5'UTR) at position 18,684-18,751 bp. Thus, TRXPT_21 is a three-exon transcript encoding DBP (380 residues, 43.3 kDa) precisely as predicted. This transcript (TRXPT_21) was also corroborated in a 3'RACE experiment. Additionally, from the 3'RACE, a splice variant of TRXPT_21 which retains the second intron leading to a 2-exon transcript was found. This transcript (TRXPT_21B), albeit longer due to retaining the second intron and possessing a short 3' UTR, encodes a truncated isoform of DBP (tDBP) because the SSC utilized by TRXPT_21, is followed shortly by STCs in the retained intron. The SSC 173 bp downstream of DBP's SSC yields tDBP (a 346 residue, 39.3 kDa product), which is in-frame of DBP but entirely contained in the second exon. TRXPT_21 and TRXPT_21B share a common TTS but TRXPT_21B as seen in our 3'-RACE data, extends 39 bp into an adenine-thymine (A-T) rich sequence before the poly-A tail sequence occur, suggesting this position (16,934bp) as the bona fide E2A TTS (**Figure 6**).

The E2B region transcripts also start with the E2-5'UTR but extend thousands of base pairs downstream to reach the TTS at 2334bp in the IM region, which is immediately followed by an A-T rich sequence (position 2323-2339bp) where polyadenylation probably occurs. Interestingly, the TTS of the E1 region (position 2,325bp) on the sense strand is also in the immediate vicinity of this A-T rich sequence, which is almost palindromic; hence it likely serves as the polyadenylation signal for both E1 and E2B/IM transcripts. The E2B transcripts, TRXPT_6 and TRXPT_7 are almost identical except for an extra splice junction at the 3'-end of TRXPT_6, making TRXPT_6 a five-exon transcript and TRXPT_7, four exons (**Figure 6**). TRXPT_7 has the CP for both classical proteins (pTP and Ad-pol) encoded in this region, of which the pTP ORF is predicted to be spliced from two exons just like in all other AdVs. The predicted splice junction of pTP is corroborated by our data; however, the full transcript is markedly longer than the predicted ORF: there are two novel non-coding 5' exons, the third exon (containing the SSC of pTP) is significantly longer than predicted, and the last exon containing the bulk of the CDS is more than triple the predicted size of pTP. The first two exons are 5'-UTRs because the SSC here is immediately followed by STCs; hence, the 5'-most SSC (position 10,995bp) of the third exon which matches the predicted SSC of pTP is utilized. The encoded product is identical to the predicted pTP protein (597 residues; 70.5 kDa). If secondary SSC (secSSC) usage is considered, with SSC at 6768bp and STC at 3430bp, the encoded product is identical to the predicted Ad-pol (polymerase) ORF (1112 residues; 129.2 kDa). TRXPT_6 differs from TRXPT_7 by containing an extra splice site at 3447-3515bp. However, the CP remains similar to that of TRXPT_7 except the Ad-pol encoded from the secSSC is a truncated isoform with a new STC resulting from the splice site.

While both TRXPT_6 and TRXPT_7 have the CP for Ad-pol with secSSC usage, in all AdVs studied, the two proteins (pTP and Ad-pol) are encoded by separate mRNAs with identical first three 5' exons and TTS, but the splice junction to the terminal exons are different. We checked for a longer splice junction between the third and fourth (terminal) exons of TRXPT_7 with our junction validation method (targeted PCR, cloning, and Sanger sequencing) and discovered a unique splice junction (10,981-7062bp) not found in our RNA-seq data. If initiated from the E2 TSS and terminated at the E2 TTS, this transcript(TRXPT_31) would encode Ad-pol exactly as predicted as its 5'-most CDS (**Figure 6**).

@@

In THEV, only one ORF (E3) was predicted in this region. However, as the E3 TU is nested in the MLTU, transcripts from the L4P (100K, 22K, 33K, and pVIII) not only overlap the E3 region transcripts entirely as seen in our RNA-seq results, but also have their TSS and TTS in practically the same locations (**Figure 7**). Therefore, we have categorized these two groups together as E3 transcripts.

We identified seven novel transcripts here (`r paste0("TRXPT_", c(22:27, 29), collapse = ", ")`) from our RNA-seq data, all originating from two distinct TSSs -- one corresponding to the TSS of the L4P (position 18,230bp) and the other at 18,727bp corresponding the E3 promoter (E3P). These E3 transcripts collectively have the CP for several predicted THEV ORFs: 100K, 22K, 33K, pVIII, E3, Fiber (IV), and ORF7 belonging to the MLTU. But some CDSs are nonidentical due to unpredicted splicing or the use of an in-frame upstream SSC. For instance, 33K is one of the few THEV ORFs predicted to be spliced from two exons; however, we discovered a significantly longer four-exon ORF (e33K) on TRXPT_24 that contains it almost entirely. The first two exons of e33K were not predicted but the last two match the predicted exons and the CDS is in-frame, but the first 20bp of the predicted 33K (including the SSC at 20,142bp) is spliced out as part of the second intron of TRXPT_24. Thus, the bona fide 33K (e33K) is a 19.8 kDa, 171 residue protein spanning four exons instead of the predicted 120 aa protein. TRXPT_24 also has the CP for the ORFs, pVIII and a longer variant of E3 (eE3; starting from an in-frame upstream SSC) if we consider downstream SSC usage. TRXPT_29 is the shortest transcript in this TU. It is a two-exon transcript, both exons comprising the CDS. The product of TRXPT_29 is a novel 73 residue protein (8.3KI) sharing the SSC of e33K but with a unique STC. TRXPT_23 being spliced identically as TRXPT_29 also encodes 8.3KI from its first SSC. Similarly, TRXPT_22 also encodes a 73 aa novel protein (8.3KII) from its first SSC that shares over 80% similarity with 8.3KI, but it differs from 8.3KI at the C-terminus. Considering downstream SSC usage, both TRXPT_22 and TRXPT_23 can encode pVIII and eE3 in that order, but TRXPT_23 being longer, has the CP for the Fiber ORF also. As the splice junctions of TRXPT_22, TRXPT_23, TRXPT_24, and TRXPT_29 essentially share the same genomic space, their validation was done with a single primer pair and they were differentiated from each other by cloning and Sanger sequencing (**Supplementary PCR methods**).

>>>>

Myriads of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs -- specifically human AdVs. Thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV genes are transcribed in a temporal manner; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late unit (MLTU or major late promoter [MLP] region), which generates five families of late mRNAs (L1-L5) based on the polyadenylation site. An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host cell modulating proteins, primarily involved in DNA replication or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins that act as capsid proteins, promote virion assembly, and direct genome packaging. The immediate early gene E1A is expressed first, followed by the the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. Noteworthily, the MLP shows basal transcriptional activity during early infection (before DNA replication), with a comparable efficiency to other early viral promoters, but reaches its maximal activity during late infection (after DNA replication). However, during early infection the repertoire of late transcripts from the MLP is restricted until late infection [@Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. For instance, the MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader, TPL) [@Donovan2020; @Zhao2014]. There is also an alternate 5' three non-coding exons present in varying amounts on a subset of MLTU mRNAs (known as the x-, y- and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5) based on polyadenylation site. Each transcription unit (TU) contains its own promoter driving the expression of all the array of mRNA transcripts produced via alternative splicing in the unit [@Zhao2014; @Donovan2020; @Guimet2016]. The promoters are activated at different phases of the infection by proteins from previously activated TUs. Paradoxically, the early-to-late phase transition during infection requires the L4 genes, 22K and 33K, which should only be available after the transition. However, a promoter in the L4 region (L4P) that directs the expression of these two proteins independent of the MLP was found, resolving the paradox [@Morris2010; @Guimet2016; @Donovan2020]. During translation of AdV mRNA, recent studies strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020; @Westergren2021].

High throughput sequencing methods have facilitated the discovery of many novel transcribed regions and splicing isoforms. It is also a very powerful tool to study alternative splicing under different conditions at an unparalleled depth [@Djebali2012; @Zhao2014; Westergren2021]. In this paper, a paired-end deep sequencing experiment was performed to characterize for the first time the transcriptome of THEV (VAS vaccine strain) during different phases of the infection, yielding the first THEV splicing map. Our paired-end sequencing allowed for reading `r seq_len` bp long high quality (mean Phred Score of 36) sequences from each end of cDNA fragments, which were mapped to the genome of THEV.
\newpage

@@

**Overview of sequencing data and analysis pipeline outputs**  
A previous study by Aboezz *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*Meleagris gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (**Table 1**, **Figure 2a**). From the mapped reads, we identified a total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junction and mapping reads to the THEV genome over time denotes an active infection, and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (**Figure 2b**).

@@

We validated the splice junctions in all transcripts by PCR amplification of viral cDNA, cloning, and Sanger sequencing (**Supplementary PCR methods**). During validation, we identified 5 additional transcripts some of which were further validated by 3' Rapid Amplification of cDNA Ends (3'RACE) data. The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (<http://www.ncbi.nlm.nih.gov/geo>) under accession number GSE254416.

@@

We also analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. We found that most splice donor-acceptor sequences were unsurprisingly the canonical GT-AG nucleotides. However, the splice acceptor-donor pairing became less specific over time, such that all combinations of nucleotide pairs were eventually detected (**Figure 5**)

@@

Only two ORFs (ORF1 [sialidase] and Hyd) are predicted in this region; however, we discovered `r as.character(english(nrow(filter(spliced_gtf, region == "E1"))))` novel transcripts in this region, which collectively contain `r nrow(reg_e1_brkdown)` unique splice junctions (**Figure 6**). Most of the ORFs of the novel transcripts are distinct from the predicted ORFs, but they all have the coding potential (CP) for the predicted Hyd protein as the 3'-most coding sequence (CDS) if secondary start codon usage is considered as reported for other AdVs [@Donovan2020; @Zhao2014]. The 5'-most CDS of TRXPT_1 is multi-exonic, encoding a novel 17.9 kilodalton (kDa), 160 residue [amino acids (aa)] protein (ORF9). From its 5'-most start codon (SSC), TRXPT_2 encodes the largest protein in this region -- a 64.3 kDa, 580 aa protein (ORF10) with the same SSC as ORF9 (position 211bp). ORF10 spans almost the entire predicted ORF1 and Hyd, coming short in two regards: it is spliced from 1655bp to 1964bp (ORF1's C-terminus, including the stop codon), and it's stop codon (STC; position 2312) is 13 bp short of Hyd's STC. However, it has an SSC 102 bp upstream and in-frame with ORF1's predicted SSC. Thus, ORF10 shares substantial protein sequence similarity with ORF1 but not with Hyd, as the SSC of Hyd is not in-frame. Without its splice site removing the ORF1 STC, TRXPT_2 would encode a longer variant of ORF1, starting from an upstream SSC. TRXPT_3 is almost identical to TRXPT_1, except for the lack of TRXPT_1's second exon. Our RNA-seq data shows that all E1 transcripts share the same transcription termination site (TTS; at position 2325bp). However, TRXPT_3 and TRXPT_4 seem to have transcription start sites (TSS) downstream of the TSS of TRXPT_1 and TRXPT_2 (E1 TSS; position: 54bp). Given that studies in MAdVs show that E1 mRNAs share not only a common TTS but also the TSS, and only differ from each other regarding the internal splicing [@Zhao2014], it is likely that TRXPT_3 and TRXPT_4 are incomplete, and their actual TSS just like the TTS are identical for all E1 transcripts. Regardless of the TSS considered for TRXPT_3, the coding potential (CP) remains unaffected. Its 5'-most CDS, beginning at 1965bp and sharing the same STC as ORF9, produces a 13.1 kDa, 115 residue protein (ORF4). ORF4 was predicted in an earlier study [@Pitcovski1998] but was excluded in later studies [@Beach2009a; @Davison2003]; however, our data suggests it is a bona fide ORF. Unlike TRXPT_3, the CP of TRXPT_4 is affected by the TSS considered; if we consider its unmodified TSS, then its CP is the same as TRXPT_3 (ORF4 as the first CDS and Hyd as second CDS using the secondary SSC). However, if we assume that TRXPT_4 uses the E1 TSS, then the 5'-most CDS is a distinct, novel, multi-exonic 15.9 kDa, 143 aa protein (ORF11) with the same SSC as ORF9 and ORF10 but with a unique STC. The splice junctions of all transcripts in this region (except the junction for TRXPT_4) were validated by cloning of viral cDNA and Sanger sequencing (**Supplementary PCR methods**).

@@

**Early Region 2 (E2) and Intermediate Region (IM) transcripts**     
The E2 TU expressed on the anti-sense strand is subdivided into E2A and E2B and encodes three classical AdV proteins --  pTP and Ad-pol (E2B proteins), and DBP (E2A protein) -- essential for genome replication [@Donovan2020; @Zhao2014]. Unlike MAdV where two promoters (E2-early and E2-late) are known [@Donovan2020], we discovered only a single TSS (E2 TSS; 18,751bp) from which both E2A and E2B transcription is initiated. However, similar to MAdVs, E2A and E2B transcripts have distinct TTSs, and the E2B transcripts share the TTS of the IVa2 transcript of the IM region [@Donovan2020; @Zhao2014] (**Figure 7**).

The E2A ORF, DBP is one of three THEV ORFs predicted to be spliced from two exons. The corresponding transcript (TRXPT_21) found in our data matches this predicted splice junction precisely but with a non-coding additional exon at the 5'-end (E2-5'UTR) at position 18,684-18,751 bp. Thus, TRXPT_21 is a three-exon transcript encoding DBP (380 residues, 43.3 kDa) precisely as predicted. This transcript (TRXPT_21) was also corroborated in a 3'RACE experiment. Additionally, from the 3'RACE, a splice variant of TRXPT_21 which retains the second intron leading to a 2-exon transcript was found. This transcript (TRXPT_21B), albeit longer due to retaining the second intron and possessing a short 3' UTR, encodes a truncated isoform of DBP (tDBP) because the SSC utilized by TRXPT_21, is followed shortly by STCs in the retained intron. The SSC 173 bp downstream of DBP's SSC yields tDBP (a 346 residue, 39.3 kDa product), which is in-frame of DBP but entirely contained in the second exon. TRXPT_21 and TRXPT_21B share a common TTS but TRXPT_21B as seen in our 3'-RACE data, extends 39 bp into an adenine-thymine (A-T) rich sequence before the poly-A tail sequence occur, suggesting this position (16,934bp) as the bona fide E2A TTS (**Figure 7**).

The E2B region transcripts also start with the E2-5'UTR but extend thousands of base pairs downstream to reach the TTS at 2334bp in the IM region, which is immediately followed by an A-T rich sequence (position 2323-2339bp) where polyadenylation probably occurs. Interestingly, the TTS of the E1 region (position 2,325bp) on the sense strand is also in the immediate vicinity of this A-T rich sequence, which is almost palindromic; hence it likely serves as the polyadenylation signal for both E1 and E2B/IM transcripts. The E2B transcripts, TRXPT_6 and TRXPT_7 are almost identical except for an extra splice junction at the 3'-end of TRXPT_6, making TRXPT_6 a five-exon transcript and TRXPT_7, four exons (**Figure 7**). TRXPT_7 has the CP for both classical proteins (pTP and Ad-pol) encoded in this region, of which the pTP ORF is predicted to be spliced from two exons just like in all other AdVs. The predicted splice junction of pTP is corroborated by our data; however, the full transcript is markedly longer than the predicted ORF: there are two novel non-coding 5' exons, the third exon (containing the SSC of pTP) is significantly longer than predicted, and the last exon containing the bulk of the CDS is more than triple the predicted size of pTP. The first two exons are 5'-UTRs because the SSC here is immediately followed by STCs; hence, the 5'-most SSC (position 10,995bp) of the third exon which matches the predicted SSC of pTP is utilized. The encoded product is identical to the predicted pTP protein (597 residues; 70.5 kDa). If secondary SSC (secSSC) usage is considered, with SSC at 6768bp and STC at 3430bp, the encoded product is identical to the predicted Ad-pol (polymerase) ORF (1112 residues; 129.2 kDa). TRXPT_6 differs from TRXPT_7 by containing an extra splice site at 3447-3515bp. However, the CP remains similar to that of TRXPT_7 except the Ad-pol encoded from the secSSC is a truncated isoform with a new STC resulting from the splice site.

While both TRXPT_6 and TRXPT_7 have the CP for Ad-pol with secSSC usage, in all AdVs studied, the two proteins (pTP and Ad-pol) are encoded by separate mRNAs with identical first three 5' exons and TTS, but the splice junction to the terminal exons are different. We checked for a longer splice junction between the third and fourth (terminal) exons of TRXPT_7 with our junction validation method (targeted PCR, cloning, and Sanger sequencing) and discovered a unique splice junction (10,981-7062bp) not found in our RNA-seq data. If initiated from the E2 TSS and terminated at the E2 TTS, this transcript(TRXPT_31) would encode Ad-pol exactly as predicted as its 5'-most CDS (**Figure 7**).

@@

In THEV, only one ORF (E3) was predicted in this region. However, as the E3 TU is nested in the MLTU, transcripts from the L4P (100K, 22K, 33K, and pVIII) not only overlap the E3 region transcripts entirely as seen in our RNA-seq results, but also have their TSS and TTS in practically the same locations (**Figure 8**). Therefore, we have categorized these two groups together as E3 transcripts.

We identified seven novel transcripts here (`r paste0("TRXPT_", c(22:27, 29), collapse = ", ")`) from our RNA-seq data, all originating from two distinct TSSs -- one corresponding to the TSS of the L4P (position 18,230bp) and the other at 18,727bp corresponding the E3 promoter (E3P). These E3 transcripts collectively have the CP for several predicted THEV ORFs: 100K, 22K, 33K, pVIII, E3, Fiber (IV), and ORF7 belonging to the MLTU. But some CDSs are nonidentical due to unknown splice sites or the use of an in-frame upstream SSC. For instance, 33K is one of the few THEV ORFs predicted to be spliced from two exons; however, we discovered a significantly longer four-exon ORF (e33K) on TRXPT_24 that contains it almost entirely. The first two exons of e33K were not predicted but the last two match the predicted exons and the CDS is in-frame, but the first 20bp of the predicted 33K (including the SSC at 20,142bp) is spliced out as part of the second intron of TRXPT_24. Thus, the bona fide 33K (e33K) is a 19.8 kDa, 171 residue protein spanning four exons instead of the predicted 120 aa protein. TRXPT_24 also has the CP for the ORFs, pVIII and a longer variant of E3 (eE3; starting from an in-frame upstream SSC) if we consider downstream SSC usage. TRXPT_29 is the shortest transcript in this TU. It is a two-exon transcript, both exons comprising the CDS. The product of TRXPT_29 is a novel 73 residue protein (8.3KI) sharing the SSC of e33K but with a unique STC. TRXPT_23 being spliced identically as TRXPT_29 also encodes 8.3KI from its first SSC. Similarly, TRXPT_22 also encodes a 73 aa novel protein (8.3KII) from its first SSC that shares over 80% similarity with 8.3KI, but it differs from 8.3KI at the C-terminus. Considering downstream SSC usage, both TRXPT_22 and TRXPT_23 can encode pVIII and eE3 in that order, but TRXPT_23 being longer, has the CP for the Fiber ORF also. As the splice junctions of TRXPT_22, TRXPT_23, TRXPT_24, and TRXPT_29 essentially share the same genomic space, their validation was done with a single primer pair and they were differentiated from each other by cloning and Sanger sequencing (**Supplementary PCR methods**).

In addition to corroborating the splice junctions for the aforementioned transcripts, the Sanger sequencing results also showed another splice  variant undetected in our RNA-seq transcriptome. This was a three-exon transcript (TRXPT_30) with its first and last exons spliced identically as TRXPT_23, but which also has the second exon of TRXPT_24 (**Figure 8**). The first CDS on TRXPT_30 spans all three exons, producing a novel 140 residue, 15.7kDa protein (e22K). Interestingly, the last 81 C-terminus residues of e22K are identical to 22K (89 residues), which is a single-exon ORF predicted to use the same SSC as 33K (20,142bp). Just as seen for 33K, all the transcripts in this region exclude the first 20bp of 22K (including the SSC) as part of their introns; therefore, the first 7 residues of 22K are lacking in e22K due to splicing. Hence, we may consider e22K as a long variant of the predicted 22K ORF. Albeit the TSS and TTS of TRXPT_30 was not seen, we presume that they are similar to TRXPT_23, in which case it would also have the downstream CP of TRXPT_23. TRXPT_25 is the largest transcript in the TU. It also utilizes the L4P TSS but has a distinct TTS. It is a two-exon transcript, encoding a novel protein (t100K; 543 residues), which is a shorter isoform of the predicted 100K ORF. Considering secSSC usage on this transcript yields the predicted 22K ORF precisely. It also has the CP for pVIII and eE3 in that order. Furthermore, during the validation of TRXPT_25's splice junction using primers that span its junction (18350-18717bp), we noticed a DNA band that corresponds to the full unspliced sequence (**Supplementary PCR methods**). As TRXPT_25 only falls short of encoding the complete predicted 100K protein due to its splice junction, this band (which we cloned and validated by Sanger sequencing) suggests that the predicted 100K is indeed expressed. This transcript (TRXPT_25B) although not seen in full, likely shares the same TSS and TTS as TRXPT_25. Lastly, TRXPT_26 and TRXPT_27 both originate from the E3 TSS but have distinct TTSs. TRXPT_26 is a three-exon transcript but the first two are UTRs. It encodes pVIII as the 5'-most ORF and has the CP for eE3 and Fiber in that order. TRXPT_27 on the other hand, is only a two-exon transcript but similar to TRXPT_26, only the terminal exon contains the CDSs. It encodes Fiber as the 5'-most ORF, and ORF7 downstream with secSSC usage. TRXPT_13, which is an L4 transcript that uses the MLP TSS is discussed under the MLTU transcripts.

**Early Region 4 (E4) transcripts**   
This TU is found at the tail-end (3'-end) of the genome  and expressed from the anti-sense strand. Based on nucleotide position, ORF7 and ORF8 were predicted in this region [@Davison2003]; however, as ORF7 is neither on the same strand as ORF8 nor transcribed from a promoter in the E4 region, only ORF8 can legitimately be classified as a transcript in this TU. This is corroborated by our RNA-seq data, as only one transcript was identified in this region on the anti-sense strand (**Figure 9**). The transcript (TRXPT_28) spans 25192-26247bp and is spliced at 25701-26055bp, making a two-exon transcript. The second exon fully matches the predicted ORF8 with 12 extra base pairs at the 3'-end. However, there is a SSC in the first exon at position 26246bp (192bp upstream of the predicted SSC). The encoded protein from this SSC is in-frame with the predicted SSC found in the second exon; hence, we consider this protein (eORF8 -- 26.4 kDa, 229 aa), a longer isoform of the predicted ORF8, the genuinely expressed ORF with an identical C-terminus to the predicted ORF8 protein.
